\documentclass[11pt,a4paper]{article}
\usepackage{amsmath, amssymb}
\usepackage{tikz}
\usepackage[margin = 1in]{geometry}
\usepackage{xcolor}
\usepackage{float}
\usepackage{incfig}
\usepackage[colorlinks=true, linkcolor=blue]{hyperref}

\title{Diary 2026}
\author{Mezbah}
\date{}

\begin{document}
\maketitle
\tableofcontents 

\newpage

\section{April, 2026}

\subsection{2026-04-16}
I have been meaning to have a diary of sort, anything where I can consistently keep a record of what I do, for a long long time. Though I used some paper notebooks and digital systems, I was never quite satisfied with tracking my progresses and recording my thoughts. The reason is that I was never really organized, and every time I started something, it didn't go anywhere, because I probably stopped mid-way somewhere, or something like that. Things were repetitve, and I never really particularly managed to produce anything. I wish to be creative, but see, one must make that creation come to life. I really truly haven't created anything yet. I realize that whatever system that I may wish to use, the system need to have a few criteria. It has to have a certain organizational structure, longevity, and something that I can look back and review, all quite consistently. Finally I have decided to use latex instead, and use just one tex file for one year, and another one for another year. This way I will have a consistent record of things that I would do. I am thinking of using this diary for all purposes, both scientific and personal. 

I shall also use this diary to track progresses in the projects that I am working on. Indeed I should use it almost on a regular basis, simply writing along. With that being said, I am currently engaged with the following projects. 

\begin{itemize}
	\item Master's thesis: on techniques of quantum field theory in the context of cosmology. 
	
	\item Quantum computing: with Apu and Alvi. 
\end{itemize}
Pretty much this two stuff are what I am seriously working on. Other than this I have some stuff to seriously investigate in philosophy or psychology. That is, some works of Carl Jung, some alchemical texts, Charles Sherrington's book on 'Man on his nature', and to reread Schrodinger's books e.g. "Mind and Matter", etc. 

I held a weekly session with Sabiha tonight after a long time. This went really nice. I have to keep this up consistently and I am sure something very nice will come out of it. And even if nothing comes out of it, the time spent together itself will be more than worth its while. 




\subsection{2026-04-17}
In today's entry I want to concisely write down a few things that I have been doing for the past few weeks, simply for the sake keeping a record of it. I have been tinkering with the computer for a while now and did a number of interesting and useful things. Some days ago I set up a neovim + inkscape + latex workflow for seamlessly adding figures while writing latex on neovim without having to leave the editor, basically making the process automatic, and I did this for both native-Windows and Linux. Then I put the whole thing on GitHub so that I can set it up in any other computer quickly without having to redo everything. And since it is on GitHub, the way it works is that anyone can actually set up my workflow in their system, if they so wish for whatever reason. I also created a neovim plugin that one can install in their neovim editor and integrate inkscape and latex. I just noticed that a week ago some guy, who seems to be very active on GitHub, found my GitHub repositories and gave a star. Nice!  

And another interesting and really useful thing that I have found is that one can use the Wolfram Engine on one's local machine, but without the Wolfram Notebook, though one gets the full computational power of Mathematica. So I did set that up. You can either run the Wolframscript on the commandline or you can use the Wolfram Language kernel on Jupyter Lab. I set up the latter because you can render the equations in a more readable form there. So now I have the computational superpower of Mathematica in my local computer! How cool is that! 

I am suddenly reminded of the matter that one of the things that I wish to do is write up my findings and understanding of love in an organized and publishable way; I say publishable only because it would force me to deepen my understanding and organize the ideas properly. Over the last two years at least I have often actively and intensely sought a deeper understanding of the phenomenon of love and I learned a tremendous amount of stuff, from books, music, songs, academic papers, and of course my own experiences and those of others, etc. It led me to many interesting directions, perhaps the most interesting that I found was the game theory, which you can use to mathematically model social dynamics. Indeed John Nash got his Nobel Prize on that kind of work. As you can imagine, a deeper understanding of love requires a deeper understanding of the principles of social interactions at a sufficiently broad level. There was another book that I came across, titled 'The denial of death', and it was written thoroughly from psychological point of view and seemed to contain an overcomplicated and a bit of a vague analysis of the human condition, though it was rather intriguing and often thought-provoking. However, I wasn't sure how much I should have taken those ideas seriously, but I like how it attempted to tackle these difficult human problems. One other notable book that I should mention is Bertrand Russell's on marriage. At any rate I learned a lot of stuff, I deepened my understanding to some extent, but the mystery is left unsolved. I am not sure anyone actually has solved it in an entirely satisfactory way. But I believe one can reach very good partial understanding, if not full, and for practical purposes that should be enough. Before I carry out this task, however, I first wish to do science. So I am devoting my energies into that at present. 

\subsection{2026-04-18}
As for my master's thesis I have the following tasks at hand that I am occupied with and need to finish as soon as possible. 
\begin{itemize}
	\item Assigned: explain the idea of dimensional transmutation to Arshad sir. 
	
	\item Work out everything in Brandenberger. 
	
	\item On the monopolo problem: Rajantie's paper: \href{https://arxiv.org/pdf/1704.04801}{Thermal Schwinger pair production at arbitrary coupling}
\end{itemize}
What I am currently engaged with is the first task. Since I feel my bakcground knowledge in QFT is still substantially lacking, and even when not knowledge then certainly the understanding, I decided to read through Schwartz's book on QFT from the beginning all the way upto all the chapters on renormalization. But that would require quite a bit of time, and since I am running short in time, I am reading the following chapters: 8, 9, 13-24, and chapter 34, the background fields, which is the most relevant to what I am working on. Lots of work to do. At present I am now on chapter 9, the scalar QED. 


\subsection{2026-04-19}
I couldn't finish last day's work that I set out to do. Need to plan more efficiently next time. I was thinking that people who actually are very productive and get a lot of work done, it might not be so much the fact that they work all the time or that they are way smarter, though they certainly work a lot and very hard and certainly very smart, but that they have better strategy or a more efficient system for getting things done. A lousy system of work can slow you down by a lot. Anyway, the point is, I need to be more efficient with my work style and craft it perfectly so that I am working at my maximum capacity while not burning myself out, so that I can keep doing interesting work successfully for as long as possible.

Before beginning today's work I want to enlist a few things that I came across yesterday, so that I can explore this stuff later. I came across the following papers that I wish to understand. 

\begin{itemize}
	\item \href{https://sci-hub.usualwant.com/10.1103/PhysRevD.46.4421}{Quantum field theory in spaces with closed time curves (David G. Boulware)}

	\item \href{https://sci-hub.usualwant.com/10.1103/physrevd.44.3197}{Quantum mechanics near closed timelike lines (David Deutsch)}
\end{itemize}

Another thing that has for some time caught my attention is the programming language called Lisp. I heard a lot of good things about it, some calls it \href{https://www.youtube.com/watch?v=OyfBQmvr2Hc&list=PLISEtDmihMo1rzX3l50vAo61bqpK5tF5x&index=1}{the most beautiful program ever written}, so I wish to learn it, but I still don't have any specific use case for it so I am not sure if I should invest the time in learning it yet, though I am pretty sure once I learn it I will be able to use it for many things, since it's a general-purpose programming language. One thing that you can do with Lisp is configure the Emacs editor (see this lecture: \href{https://www.youtube.com/watch?v=2z-YBsd5snY&list=PLISEtDmihMo1rzX3l50vAo61bqpK5tF5x&index=3}{An introduction of Emacs Lisp}). Since I am using neovim as my editor, I don't need that now. I have in mind two areas where I wish to use it: one, to increase productivity, and two, applications in physics. I came across the following comment in a \href{https://www.reddit.com/r/lisp/comments/utuj9i/lisp_for_quantum_simulation/}{reddit thread}: 
\begin{quote}
	"Lisp lacks scientific computing libraries. Lisp does not have a lot of scientific computing users. Python, Julia, Matlab, and Mathematica have sucked up 95\% of scientific programmers, and none of them will be terribly interested in writing Lisp. I think Julia (which I don't advocate using) demonstrated what Lisp had the capacity to offer, but nobody stepped up to the plate to do.

	So linear algebra, differential equation solvers, optimizers, integrators, special functions, plotting, statistical routines, root finders, geometry routines, arbitrary precision arithmetic, etc. etc. collectively are in a sorry state in Lisp. There's no cohesive answer for how to do any of that. There are some small things here and there that solve certain subproblems in those categories, but you'll never rest easy knowing there's a community behind it all supporting it."
\end{quote}
Hmm, so clearly not going to need it for scientific computing. So that leaves me only with the Emacs, if I ever use it. I need to make it a rule to only learn things that I will have serious use-case for. Otherwise I would end up wasting time learning things that will not be very necessary other than feeding the curiosity. 

\sloppy My present task in physics is understanding renormalization. Though I am studying Schwartz's book on QFT, I will enlist a few other sources which I wish to read. 

\begin{itemize}
	\item Peskin and Schroder 
	\item Ashok Das: chapter 16, 17. 
	\item V. P. Nair: chapter 9 
	\item Advanced QFT lecture notes, part III of Mathematical Tripos, Cambridge (Linus Chang): chapter 3 to the end. 
	\item Also see the following UT Austin lecture notes: \href{https://web2.ph.utexas.edu/~vadim/Classes/2024f-qft/RG.pdf}{renormalization group techniques},  \href{https://web2.ph.utexas.edu/~vadim/Classes/2024f-qft/ms.pdf}{minimal subtraction} and \href{https://web2.ph.utexas.edu/~vadim/Classes/2024f-qft/RGT.pdf}{renormalizability of gauge theories}. 
\end{itemize}

Needless to say, I am now feeling that I should have started exploring a lot earlier. Also another thing that's coming out is the inadequacy of my previous learning of QFT. Even though I did follow the calculations and did some problems, in no way did I develop a broad understanding of the subject, though I believe I should have been able to. I just needed to be more strategic and more efficient and basically should have studied everything from a number of books. That's the only way to have a broad view and understanding of a subject, you must have a broad picture and see how everything fits together, while trying your hands on simple problems that would illustrate the ideas. Okay, that's not the only way. Another way to develop broad understanding is to have detail conversations with someone who's already an expert on the subject-matter, or a certain topic. You can pick things up very fast simply from such very productive discussions or conversations. 





\subsection{2026-04-20}
I was studying the Schwinger-Dyson equations and the Ward-Takahashi identity, and I am quite fascinated by both. The graphical representation of the Schwinger-Dyson differential equation is given in Itzykson and Zuber, and I don't understand it. 

I looked up on the web and I came across a few stuff that I wish to look into later. I will enlist those here. I found two lecture notes of some professor from University of Texus, Austin: Ward-Takahashi identities: \href{https://web2.ph.utexas.edu/~vadim/Classes/2024f-qft/WTI1.pdf}{current conservation} and \href{https://web2.ph.utexas.edu/~vadim/Classes/2024f-qft/WTI2.pdf}{diagrammatic proof}. At present this looks beyond me and I would need some solid time to sit through these to understand how it works. Also look at his whole QFT1 and QFT2 \href{https://web2.ph.utexas.edu/~vadim/Classes/2024f-qft/notes.html}{course website}, and indeed all \href{https://web2.ph.utexas.edu/~vadim/class_pages.html}{his classes}, which is excellent. I included in the previous entry his lecture notes on renormalization, which looked nice. And I came across a few papers on Ward-Takahashi identities: 
\begin{itemize}
	\item A very recent paper on Physcial Review, published this year: \href{https://journals.aps.org/prd/abstract/10.1103/jm5x-7d83}{Ward-Takahashi identity in the light-front formalism for a bound state of fermions}
	\item \href{https://arxiv.org/pdf/1608.08134}{The full Ward-Takahashi identity for colored tensor models}
\end{itemize}

I came across another paper which is not related to renormalization, but on the strong CP problem and axions, which I wish to learn more about. There are some unsolved mysteries there. 
\begin{itemize}
	\item \href{https://arxiv.org/pdf/hep-ph/0607268}{The Strong CP Problem and Axions (R. D. Peccei)}
\end{itemize}



\subsection{2026-04-21}
I didn't have much sleep last night and it's now 11:49 am and I haven't yet gotten to study. I have been browsing internet pretty much randomly since morning today. I want to fix a certain amount of time to browse the internet and read various blog posts and stuff. I shall enlist a number of websites on physics or mathematics. 

\subsubsection{List of some websites}
The following list does not follow any order at all. They are listed randomly.
\begin{itemize}
	\item \href{https://www.damtp.cam.ac.uk/user/tong/}{David Tong}
	\item \href{https://www.damtp.cam.ac.uk/user/dbs26/teaching.html}{David Skinner}: Cambridge physics lecture notes
	\item \href{https://scottaaronson.blog/}{Scott Aaranson}
	\item \href{https://nafizishtiaque.com/?i=1}{Nafiz Ishtiaque}
	\item \href{https://terrytao.wordpress.com/}{Terry Tao}
	\item \href{https://www.dpmms.cam.ac.uk/~wtg10/}{Timothy Gowers}
	\item \href{https://www.math.columbia.edu/~woit/wordpress/}{Peter Woit's Not Even Wrong}
	\item \href{https://webspace.science.uu.nl/~hooft101/}{Gerard 't Hooft}
	\item \href{https://www.cumrunvafa.org/}{Cumrun Vafa}
	\item \href{https://www.preposterousuniverse.com/blog/}{Sean Carroll}
\end{itemize}
And 
\begin{itemize}
	\item \href{https://www.quantamagazine.org/}{Quanta magazine}
	\item \href{https://quantumfrontiers.com/}{Quantum frontiers}
	\item \href{https://quantarei.wordpress.com/}{quanta rei}
	\item \href{https://blog.qutech.nl/}{Bits of quantum}
	\item \href{https://golem.ph.utexas.edu/category/}{The n-Category Cafe}
\end{itemize}
While I have enlisted some webpages to look up, there is a virtue to exercise here. There are too much stuff out there, indeed a lot of interesting things. But the brain can only absorb so much. It is important that you take in information in bite-sized chunks and never more than that. This is actually very very important to keep in mind, otherwise it is very very easy to keep reading random stuff while getting absolutely nothing out of it, and no one would want that. Anyways, I need to go have breakfast, reschedule my Google Calendar, and get to work. I am to read up on renormalization today from Schwartz' QFT book. I read up to the Casimir effect last night, how different regulators give the same answer for the Casimir force. I shall start with regulator independence proof today. 

Even though I am dealing with ideas that are quite technical and requires mathematical expressions, it is vital that I also develop the ability to explain them without any equations. I should be able to say what is happenening and how, and even how a certain calculation goes, all of which in simple plain words.

Suppose we have a scalar field theory and we calculate the scattering amplitude for the $s$-channel diagram. We introduce a cutoff to calculate the scattering amplitude, $\mathcal{M}(s) $, where $s$ is for s-channel. This amplitude is an inifinite quantity. But the difference between $\mathcal{M}(s_1)$ and $\mathcal{M}(s_2)$ at two different scales will be a finite quantity. And we use this fact to renormalize the coupling constant $\lambda$. The way we do it is that we define a renromalized coupling $\lambda _R$ as the value of the amplitude at some scale. This value is by definition something observable and finite. Then we can express $\mathcal{M}(s)$ in terms of $\lambda_R$ for any $s$ which will be finite order by order in perturbation theory. This will be a physical prediction - one observable in terms of another which we know. Note that tree-level graphs only give rational polynomials in momenta and couplings, never logarithms. Whereas the logarithmic behavior is a characteristic of loop effects. 

It's 19:20 now, evening. So far today I have finished reading the chapter 15, the first chapter on renormalization in Schwartz's book, and the Appendix B: regularization, and started with chapter 16. I think I have a bit of a better understanding of this stuff, though not yet at a working level. I should have studied these things properly so long ago though, and by this time should have been a master. 

22:53 Apu just called me on phone, since there is no electricity, and was telling me that Arshad sir took an exceedingly long class today, starting from 10 in the morning to 3 in the afternoon with a little break in between. Having taught us the path integral approach last year, he realized that starting with the path integral formalism was a mistake and so this time he will teach the canonical quantization method instead. Well, that's how the standard treatment goes though. Generally you first learn the canonical approach and then you learn the path integral method, though from what I can tell Wheeler and A. Zee seemed to prefer the path integral method. Apu also recounted some other stories, one of which I will record here - some girl in the class asked Arshad sir what $\sigma _3$ was. He replied that it was a Pauli matrix. Then the girl asked, what is a Pauli matrix? Lol. I am sure Arshad sir completely lost it at that moment. He called over the chairman of the department right away in the class to ask what sort of admission test they took that it didn't filter out people who don't know what the Pauli matrices are. That's a totally valid crashout though. Who gets into a physics grad course without knowing Pauli matrices? 


\subsection{2026-04-22}
10:30 I was just reading Schwartz's QFT book, the Appendix section B.4.3, 'Other regulators' and it's very interesting how different regulators come with different kinds of advantages and disadvantages. For instance, the dimensional regularization has the advantages that it respects gauge invariance, can regulate IR or UV divergences, new fields need not be introduced (like say in Pauli-Villars), etc, and disadvantages of being unphysical and unintuitive. Whereas the hard-cutoff regulator breaks not only the Lorentz invariance but every symmetry of the theory, but has the advantage of being the most intuitive regularization procedure. 

10:56 I was just studying and I recalled the principle that the reward of success is more work, not less, and that it's a good thing that that's how it should be. You should want to work more, not less. Of course, it should be work that is worth doing. Just wanted to write that down here.

11:11 Another interesting thing that I just came across is that the Pauli-Villars regularization method can be viewed as a systematic implementation of the derivative method, due to the following identity:
\[
	\int dm ^2 \frac{d}{dm ^2} \left[ \frac{1}{k ^2 - m ^2} \right] = \int dm ^2 \frac{1}{(k ^2 - m ^2) ^2}
\]

12:06 I will use this diary to keep little things in mind in an exceedingly concise form as I read something. Or I could also elaborate if need be. Here's the core idea behind renormalization in QFT: observables are finite and in principle calculable functions of other observables. Most of the confusions come from trying to express observables in terms of non-observable quantities like coupling constants in a Lagrangian. In practice what happens is the following: 
\begin{itemize}
	\item Infinities associated with high-energy divergences may appear in intermediate steps of calculations, such as in loop diagrams. 
	\item To get rid of the infinities you have to deform the theory using some regulator, which in the end must drop out of physical predictions. 
	\item Coefficients of terms in the Lagrangian are NOT observable. You can solve them in terms of the regulator and they will drop out of physical predictions. 
	\item Non-analytic behavior, like $\ln \frac{s}{s_0}$, is characteristic of loop effects. 
	\item Tree-level amplitudes are always rational polynomials in external momenta and never involve logarithms. 
\end{itemize}

Another important point is that: the short-distance deformation cannot depend on long-distance physical quantities. (See section 16.1)

22:37 Quantum electrodynamics has a Landau pole: perturbation theory breaks down at short-distances. It implies that QED is not a complete theory, for it can't say how to compute scattering amplitudes at all energies. 



\subsection{2026-04-23}
06:51 I am feeling quite sluggish after getting up today and not feeling like working at all, while also wishing to finish everything as soon as possible. There are a lot of things to cover. Last night I was describing to Sabiha once more what a terrible vice anger is and I wish to stay away from it as much as possible, for it makes me destroy things, even things that I hold close to my heart. I will keep a record of what I told her here: 
\begin{quote}
	"Anyway, what I meant to say is that since I was angry at father and since I was convinced that I'd always remain self-taught, that rendered academic pursuits unnecessary for me while allowing me to not give father what he wants, so I decided to forget the HSC and carve my own path instead. Later on I came across the story of Oliver Heaviside who always worked on his own, alone, and discovered many things. But Heaviside was a genius. I then thought that if I had any talent whatsoever to contribute to the world of physics and math, I could just as well do it working on my own while not being associated to academia.

	But since my father was just as much persistent to make sure that I get university education, I happened to get in eventually, while still holding the previous conviction of not caring about it at all, and hence I did not work for any exams in undergrad. However I did study a lot on my own though and learned things according to my own needs and curiosity. All this has put me in an odd position. The regret is that I remained in academia without trying my best, and I don't feel proud of that. I take more pride in my effort than in my results. That was the biggest destruction my anger inflicted upon me. And it was an unfair reaction to my father."
\end{quote}

I think I will sleep a little more and then get up to work. Without proper sleep I will  not be able to work energetically and won't be making progress. Sleep is absolutely essential. Yesterday was not entirely a waste, though I could do more. I studied through the vacuum polarization calculation in QED and learned the various regularization methods, though I need to do these calculations properly by hand again and be a master at doing these calculations.

08:00 I could not sleep and instead was looking through Itzykson \& Zuber's book, two things - the constant  field method of Schrodinger that you can use to obtain a bound for the photon mass and the Casimir effect calculation. This book leaves out the intermediate calculations a lot and expects the reader to supply it, which is fine, except I am probably somewhat less prepared to handle this book at present. I would need much more strong mathematical background and certainly a much more strong background in classical electrodynamics. 








\end{document}
